---
title: "Introduction to Bayesian Statistics Early Draft"
author: "Sam J. Vizcaino-Vickers"
date: "28/10/2024"
format: html
editor: visual
---

# 1. Introduction to Bayesian Models in R

This document provides an overview of Bayesian modeling in R, covering linear regression, logistic regression, Poisson regression, multilevel models, and custom settings in the `brms` package.

```{r setup, include=FALSE}
# Load necessary packages
library(devtools)
library(rstan)
library(brms)
library(tidyverse)
library(lme4)
```

## 1.1 Bayesian Linear Regression

The initial code simulates some data and fits a Bayesian linear regression model using Markov Chain Monte Carlo (MCMC) methods.

```{r}
# Simulated data
data_df1 <- data.frame(
  y = c(2.3, 1.9, 3.7, 2.5, 4.1, 5.2),
  x_1 = c(1.1, 0.8, 1.4, 1.0, 2.0, 1.5),
  x_2 = c(0.7, 1.2, 0.5, 1.1, 1.3, 0.9)
)

# Simple model
M1 <- brm(y ~ x_1 + x_2, data = data_df1)
summary(M1)
plot(M1)
```

**Explanation**:\
The `brms` function uses Bayesian MCMC to estimate the posterior distributions for each model parameter, assuming a Gaussian family for the continuous response variable `y`. Key statistics such as the Rhat are provided to assess chain convergence.

**Inference**:\
The model's output includes Rhat values, which help confirm convergence. The closer to 1, the more reliable the MCMC sampling.

## 1.2 Posterior Distribution Visualization

We visualise the posterior distributions and set up the model to show the parameter distribution ranges.

```{r}
# Plot posterior distributions
mcmc_plot(M1)
mcmc_plot(M1, type = 'hist', binwidth = 0.05)
```

**Explanation**:\
These visualizations provide insights into the parameter distributions, helping to confirm the reliability of our estimated values.

## 1.3 Modifying Priors and MCMC Settings

Custom settings allow adjusting iterations, chains, warmup steps, and priors.

```{r}
M2 <- brm(y ~ x_1 + x_2, data = data_df1,
          iter = 2500,
          warmup = 500,
          chains = 4,
          prior = set_prior('normal(0, 100)'))
summary(M2)
```

**Inference**:\
By extending iterations and customizing priors, we can refine the model’s robustness. A prior distribution of `normal(0, 100)` centers the values around zero, which can be useful for highly variable data

## 1.4 Model Comparison

We can now compare models with the `loo` method and `bayes_factor`.

```{r}
bayes_R2(M1)
bayes_R2(M2)
```

## 1.5 Logistic and Poisson Regression Models 

Bayesian methods are also applicable to logistic and Poisson regressions, allowing modeling of binary and count data, respectively.

```{r}
# Logistic regression
data_df2 <- data.frame(
  y = rbinom(100, 1, 0.5),
  x_1 = rnorm(100, mean = 170, sd = 10),
  x_2 = rnorm(100, mean = 25, sd = 5)
)
M4 <- brm(y ~ x_1 + x_2, data = data_df2, family = bernoulli())

# Poisson regression
data_df3 <- data.frame(
  y = rpois(100, exp(0.5 * x_1 + 0.3 * x_2)),
  x_1 = rnorm(100, mean = 5, sd = 2),
  x_2 = rnorm(100, mean = 10, sd = 3)
)
M5 <- brm(y ~ x_1 + x_2, data = data_df3, family = poisson())
```

## 1.6 Multilevel Models

With Bayesian multilevel models, we can add grouping structures to model individual differences in intercepts and slopes.

```{r}
# Random intercept and slope model
n_subjects <- 10
n_obs_per_subject <- 10
x_2 <- factor(rep(1:n_subjects, each = n_obs_per_subject))
x_1 <- rep(0:9, times = n_subjects)
y <- 15 + 2 * x_1 + rnorm(n_subjects * n_obs_per_subject, sd = 3)

data_df4 <- data.frame(y = y, x_1 = x_1, x_2 = x_2)
M10 <- brm(y ~ x_1 + (x_1 | x_2), data = data_df4)
```

### Conclusion

This document illustrated Bayesian modeling using `brms`, covering basic regression, advanced customizations, and multilevel models. Bayesian methods allow incorporating prior knowledge into models, making them adaptable for real-world, complex data. However, the issue at the moment is that this is a good start, but we need to manage the models more carefully and to also look at the equations we implement, the independent / dependent relationships, and complexity of the variables. Therefore, my next chapter will focus upon bayesian methods using `rstanarm` , `cmdstan` and `stan`.
